A primer to numerical simulations: The perihelion motion of Mercury

### 2017-07-05 tex/MercuryNumerical.tex
modified numerical implementation section

\usepackage{ amssymb }
\usepackage{graphicx}           % Import Pictures + PDFs
\usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
\usepackage{ulem}
\definecolor{fzjblue}{HTML}{005B81}
\definecolor{light-gray}{gray}{0.97}
\definecolor{mygreen}{rgb}{0,0.6,0}
\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
\lstset{
  language=Python,
  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\linespread{1}\footnotesize,        % the size of the fonts that are used for the code
  commentstyle=\color{mygreen},
  frame=single,
  keywordstyle=\color{blue},       % keyword style
  tabsize=4,	                   % sets default tabsize to 2 spaces
  rulecolor=\color{light-gray},
}

@@

\section{Introduction}

@@

\section{Trajectories, velocities, accelerations and Newtons second law}
\label{sec:tva}

@@

\section{Example: General Relativity and the perihelion motion of Mercury}

@@

%\end{itemize}

\section{Numerical Implementation}
\label{sec:Numerical Implementation}
In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular
  the perihelion motion of Mercury.  We select \python~\cite{} as  programming language, because \python is easy to learn, 
  intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
  how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
  is required because we explain all the steps necessary to create the simulation.  We also provide a working 
  example \cite{} that can be used as template.

To start the simulation one needs the "initial" distances and velocities, which are needed as input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).  As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.  \textit{Why is this simplified picture sufficient to describe the motion}?

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39$ km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in an appropriate unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$, where the time refers to earth days, and distances in $R_0 = 10^{10}$ m.  With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
\begin{equation}
	r_{MS}(0) = 4.6 R_0 \, , \quad
	v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
	\, .
\end{equation}

\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
\end{figure}

In \python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python module \texttt{Vpython}, which provides  an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position. However, we first need to fix the time step $\Delta t$ and use the estimate
of Eq.~(\ref{eq:check}) as a guidance {\bf Careful: Check units of dt}
\begin{lstlisting}
	# Definition of the time step
        dt = 2 * vm0 / aM / 100
\end{lstlisting}
Here the factor $1/100$ makes sure that $\Delta t$ is indeed
consistent with Eq.~(\ref{eq:check}).  {\bf We need to make sure that 'dt' is known
to the subroutines}
Now we are in the position to calculate
location and velocity of the planet at $t_0+\Delta t$ using the following comands
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" -- Don't Repeat Yourself)
\begin{lstlisting}
	# Define the function
	def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
		<...Code...>
		return vec_rM_new, vec_vM_new

	# Call the function
	vec_rM_new, vec_vM_new = do_time_step(vec_rM_old, vec_vM_old)
\end{lstlisting}
Because of \python{$\!'$}s syntax, it is necessary that the body of the function is indented relative to the definition statement {\bf I do not understand this sentence}.

Finally, we can express the evolution by a \texttt{while}-loop
\begin{lstlisting}
	t = 0
	alpha=0
	# Execute the loop as long as t < T
	while t < T:
		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
		t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha$ to zero in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration.  The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
To describe at least one full orbit, $T$ needs to be larger than $T>..$.

{\bf Here we need to write something about visualizing the trajectories and how to run the code }

\section{Tests of stability}

\section{Tasks}

\section{Dimensional analysis}

@@

\section{Discussion of Starting Values}

\textit{I am not quite sure what to add here to Christopher's discussion.}\\

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}. To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added. This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories. However it might be advisable to not start out with the correct mercury values but e.g. a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then. This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.

\begin{figure}[htb]
	\centering
	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
\end{figure}

\section{Extracting the Perihelion Motion}

To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the angle. There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets maximal, the aphelion, after $n$ turns and compare it to the original one. This can be done by using an $if$-statement in the $while$-loop.

Suppose we want to save the number of turns in the variable $counter$, the maximal number of turns in $maxcounter$ (here e.g. 10), the initial aphelion in the vector $a0$ and the final aphelion in the vector $a$. In addition to find out when the aphelion is reached, the last two positions $rlast$ and $rbeforelast$ have to be known. To set this up, before the while loop the following code needs to be included
\begin{lstlisting}
	r=r0
	rlast=r0
	counter=-1
	maxcounter=10
\end{lstlisting}
Then the following code needs to be included in the $while$-loop
\begin{lstlisting}
    rbeforelast=rlast
    rlast=r
    r=r+v*dt
    if (counter==-1): a0=rlast
    if (rbeforelast.mag<rlast.mag and rlast.mag>r.mag): counter=counter+1
    if counter==maxcounter:
    	a=rlast
    	break
\end{lstlisting}
and the condition for the $while$-loop should be set to $True$.
The angle between two vectors is given by the formula
 \begin{equation}
 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
 \end{equation}
This is readily implemented in \texttt{Vpython} via
\begin{lstlisting}
	angle=acos(dot(a0,a)/(a0.mag*a.mag))/maxcounter*180/pi
	print("angle: ", (angle))
\end{lstlisting}

As explained above the natural value for $\alpha$ would be $1$. However if the student try out this value, they will find, that there is no visual change in the trajectories and the distance between the first an the last aphelion is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$. Therefore to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large. The students can convince themselves of this fact, by plotting some angles over $\alpha$. This can be done either by hand or using another program like e.g. Excel. One such plot can be found in figure \ref{fig:AlphaAngle}. Then to get to the angle at $\alpha=1$ one would ideally employ a line of best fit, but to get the order of magnitude right it is enough to consider only two points. Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
\begin{equation}
	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
\end{equation}
so $3.63''\cdot 415\approx 150''$ per 100 earth years.
To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.

\section{Possible extensions}

@@

%\begin{table}
%\caption{\label{jlab1}Journals to which this document applies, and macros for the abbreviated journal names in {\tt iopart.cls}. Macros for other journal titles are listed in appendix\,A.}
%\footnotesize
%\begin{tabular}{@{}llll}
%\br
%Short form of journal title&Macro name&Short form of journal title&Macro name\\
%\mr
%2D Mater.&\verb"\TDM"&Mater. Res. Express&\verb"\MRE"\\
%\br
%\end{tabular}\\
%$^{a}$UK spelling is required; $^{b}$MSC classification may be used as well as PACS; $^{c}$titles of articles are required in journal references; $^{d}$Harvard-style references must be used (see section \ref{except}); $^{e}$final page numbers of articles are required in journal references.
%
%\end{table}
%\normalsize

\subsection{Acknowledgments}
Here they come

\appendix
\section{Instructions to install VPython}
\label{appendix:python}

\section{The code}
Here we should put the code

\bibliographystyle{alpha din}
\bibliography{MercuryNumerical}

>>>>

\usepackage{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

\usepackage{ amssymb }
\usepackage{graphicx}           % Import Pictures + PDFs
\usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
\usepackage{ulem}
\definecolor{fzjblue}{HTML}{005B81}
\definecolor{light-gray}{gray}{0.97}
\definecolor{semilight-gray}{gray}{0.67}
\definecolor{mygreen}{rgb}{0,0.6,0}
\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
\lstset{
  language=Python,
  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\linespread{1}\footnotesize\ttfamily,        % the size of the fonts that are used for the code
  commentstyle=\color{mygreen},
  frame=single,
  keywordstyle=\color{blue},       % keyword style
  tabsize=4,	                   % sets default tabsize to 2 spaces
  showtabs=true,
  rulecolor=\color{black},
  columns=fullflexible,
  literate={\ }{{\copyablespace}}1
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trajectories, velocities, accelerations and Newtons second law}\label{sec:tva}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Implementation}\label{sec:Numerical Implementation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%------------------------------------------
\subsection{Describing the Motion with Python}
%------------------------------------------

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python~\cite{}\todo{Citation} as  programming language, because \python is easy to learn, 
intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
is required because we explain all the steps necessary to create the simulation.  We also provide a working 
example \cite{}
%
\todo{One can additional files to the arXiv upload} that can be used as template.
%

To start the simulation one needs the "initial" distances and velocities, which are needed as input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).  As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in an appropriate unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1$d, where $d$ refers to earth days, and distances in $R_0 = 10^{10}$ m.  With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
\begin{equation}
r_{MS}(0) = 4.6 R_0 \, , \quad
v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
\, .
\end{equation}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
\end{figure}
%

In \python, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
aM  = 0.99   # in units of R0/T0**2
T   = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python module \texttt{VPython}, which provides  an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from visual import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position. However, we first need to fix the time step $\Delta t$ and use the estimate
of Eq.~(\ref{eq:check}) as a guidance \todo{\bf Careful: Check units of dt}
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / aM / 1000
\end{lstlisting}
Here the factor $1/1000$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
%
\todo{I need 1/1000 to get relatively closed orbits -- we have to crosscheck codes!}
%
Now we are in the position to calculate
location and velocity of the planet at $t_0+\Delta t$ using the following comands
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" -- Don't Repeat Yourself)
\begin{lstlisting}
# Define the function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	#<...Code...>
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = do_time_step(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
This is visualized by the long underscore in the previous block.
Furthermore, \python is an Interpreter language.
Each line of the code is execute once the Interpreter passes them.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the scope of the function.

Finally, we can express the evolution by a \texttt{while}-loop
\begin{lstlisting}
t     = 0
alpha = 0
# Execute the loop as long as t < T
while t < T:
	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha$ to zero in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration.  The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, $T$ needs to be larger than $T>88$d.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T = 10^4$ evolution steps.

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
Additional to the vector class, one has to include
\begin{lstlisting}
from visual import vector, sphere, color, curve, rate, display
\end{lstlisting}
The class sphere will represent mercury and the sun in the simulation
\begin{lstlisting}
# define the initial coordinates; M = mercury, S = sun
M = sphere( pos=vec_rM0,       radius=0.01, color=color.red   )
S = sphere( pos=vector(0,0,0), radius=0.1,  color=color.yellow)
# and the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
\end{lstlisting}
We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.

To further visualize the trajectory of mercury we add this property to our \texttt{mercury} object
\begin{lstlisting}
# add a visible trajectory to mercury
M.trajectory = curve(color=color.black)
\end{lstlisting}
Next we tell python that the objects shall be drawn from this point on
\begin{lstlisting}
# display the objects from now on
display(background=color.white)
\end{lstlisting}
and last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
\begin{lstlisting}
t     = 0
alpha = 0
# update the objects
while t < T:
	# set the frame rate: shows four earth days at once
	rate(4*1000)
	# update the drawn trajectory with the current position
	M.trajectory.append(pos=M.pos)
	# update the velocity and position
	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
	# update tim
	t = t + dt
\end{lstlisting}

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}. To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added. This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories. However it might be advisable to not start out with the correct mercury values but e.g. a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then. This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.
%
\todo{This is really to large for me to get these orbits. Also I need to run with $\alpha = 10^6$ to get such shapes...}
%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
\end{figure}
\todo{@CK: reproduce figure}

%------------------------------------------
\subsection{Running the Program}
%------------------------------------------
The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
Depending on the editor of choice, there might already be a execute functionality e.g. \textbf{bla...} for the recommended windows installation (see \ref{appendix:python}).
%
\todo{Look up the windows example...}
%
In case one uses a linux like system and wants to execute the code from a console/terminal, one has to to execute the command
\begin{center}
	\texttt{python 'address/to/python/file.py'}
\end{center}
If one already works in the file directory and has used the suggested naming scheme, this reads 
\begin{center}
	\texttt{python mercury\_orbit.py}
\end{center}
Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
For more information see \ref{appendix:python}.

%------------------------------------------
\subsection{Extracting the Perihelion Motion}
%------------------------------------------

To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the relative angle between the current mercury position and the starting position.
There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets minimal -- the definition of the perihelion\todo{why aphelion and not perihelion? I have changed the following code to the perihelion (because it is shorter). We can change it back if desired.} -- and compare the new position of the perihelion after $n$ turns to the original one.
This can be done by looking for a minimal distance of mercury to the sun within a $while$-loop.

Suppose we want to save the number of turns in the variable \texttt{turns}, the maximal number of turns in \texttt{max\_turns} (here e.g. 10), the initial perihelion is already stored in the vector \texttt{vec\_rM0} and the goal is to find the final perihelion.
In addition to find out when the perihelion is reached, the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} have to be known.
Thus, if the last vector is smaller than the previous and the current vector, one has reached the perihelion again.
The implementation can be realized as follows
\begin{lstlisting}
# set up vectors
vec_r_last = vec_rM0
turns      = 0
max_turns  = 10
# find perihelion after ten turns
while turns < max_turns:
	vec_r_before_last = vec_r_last
	vec_r_last        = M.pos
	#<...update mercury position...>
	# check if at perihelion
	if vec_r_before_last.mag > vec_r_last.mag < M.pos.mag:
		turns = turns+1
\end{lstlisting}
When the loop stops, the last perihelion vector is stored in  \texttt{vec\_r\_last}.
To compute the angle between two vectors one uses the formula
 \begin{equation}
 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# import function to resolve angle
from visual import acos, pi, dot
angle = acos(
	dot(vec_rM0, vec_r_last.mag) / (vec_rM0.mag * vec_r_last.mag)
) / maxcounter * 380 / 2 / pi
\end{lstlisting}

As explained above the natural value for $\alpha$ would be $1$.
However if the student us this value for $\alpha$, they will find that there is close to no visual change in the trajectories and the distance between the first an the last perihelions is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$.
Therefore, to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large.
The students can convince themselves of this fact, by plotting some angles over $\alpha$.
This can be done either by hand or using another program like e.g. Excel.
One such plot can be found in figure \ref{fig:AlphaAngle}.

To get to the angle at $\alpha=1$ one would ideally employ a linear regression (or ruler fit), but to get the order of magnitude right it is enough to consider only two points.
Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
\begin{equation}
	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
\end{equation}
so $3.63''\cdot 415\approx 150''$ per 100 earth years.
To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tests of stability}\label{sec:stability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dimensional analysis}\label{sec:analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tasks}\label{sec:tasks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Possible extensions}\label{sec:extensions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here they come

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Instructions to install VPython}\label{appendix:python}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mac}
\href{http://vpython.org/contents/download_mac.html}{Download VPython for Mac}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The code}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we should put the code

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bib
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{alpha din}
\bibliography{MercuryNumerical}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
